Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization

### 2023-04-27 conference-paper/main.tex
renamed files to main to be consistent

% [ ] TODO: AL-PA, Consider changing the title to "Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization"
\title{Comparing Hydrogen and Jet-A for an Ultra High Bypass Ratio Turbofan with Water Recirculation} % change
\author{Peter N. Atma\footnote{MSE Student, Department of Aerospace Engineering, AIAA Student Member}}
\author{Andrew hbt!.R. Lamkin\footnote{Ph.D.~Candidate, Department of Aerospace Engineering, AIAA Student Member}}
\author{Joaquim R.~R.~A.~Martins\footnote{Professor, Department of Aerospace Engineering, AIAA Fellow}}
\affil{University of Michigan, Ann Arbor, MI, 48109}

% ==================================================
%	OVERALL COMMENTS
% ==================================================
% 1. AL-PA, Fix the citation for Kalnay1996

@@

For measuring the efficiency of jet engines, thrust-specific fuel consumption (TSFC) is generally used since it is a represents how low the fuel burn is for a given level of thrust.
However, when comparing Jet-A and hydrogen fuels this is not such a good metric since a given mass flow rate of hydrogen has an energy content almost 3 times that of Jet-A.
Therefore, a new metric for comparing the relative efficiencies of engine running on Jet-A versus hydrogen is thrust-specific energy consumption (TSEC) which multiplies TSFC by the lower heating value (LHV) of the fuel \cite{Adler2023}.

@@

% Update with table from Andrew
% TODO: AL - PA, What is the citation key "Kalnay1996"?  It's not in the MDOLab Bib or references.bib
\begin{table}[hbt!]
    \centering
    \caption{
        N+3 engine model parameters.
        These parameters specify the engine design used for analysis in this report.
        Atmospheric parameters are presented that specify the flight conditions and humidity of the atmosphere.}
    \begin{tabular}{c c c c}
        \hline
        Parameter             & Value    & Units                 & Comments                                   \\
        \hline
        $W_{TOC}$             & 820.441  & $lbm/s$               & TOC total mass flow rate of air            \\
        $BPR_{TOC}$           & 23.945   & $-$                   & TOC bypass Ratio                           \\
        $PR_{HPC,TOC}$        & 53.633   & $-$                   & TOC overall pressure ratio                 \\
        $PR_{LPC,TOC}$        & 3.00     & $-$                   & TOC low-pressure compressor pressure ratio \\
        $PR_{fan,TOC}$        & 1.30     & $-$                   & TOC fan pressure ratio                     \\
        $F_{net,SLS}$         & 28620.84 & $lbf$                 & SLS net thrust                             \\
        $F_{net,CRZ}$         & 5510.72  & $lbf$                 & CRZ net thrust                             \\
        $T_{4,TOC}/T_{4,RTO}$ & 0.926    & $-$                   & TOC-to-RTO temperature ratio               \\
        $T_{4,RTO}$           & 3400.0   & $^{\circ}R$           & RTO combustor temperature                  \\
        $h_{TOC}$             & 0.00017  & $kg_{water}/kg_{air}$ & humidity ratio at TOC \cite{Kalnay1996}    \\
        $h_{CRZ}$             & 0.00017  & $kg_{water}/kg_{air}$ & humidity ratio at CRZ \cite{Kalnay1996}    \\
        $h_{RTO}$             & 0.009    & $kg_{water}/kg_{air}$ & humidity ratio at RTO \cite{Kalnay1996}    \\
        $h_{SLS}$             & 0.009    & $kg_{water}/kg_{air}$ & humidity ratio at SLS \cite{Kalnay1996}    \\
        $LHV_{JetA}$          & 18564.0  & $BTU/lbm$             & Lower heating value of Jet-A               \\
        $LHV_{H2}$            & 51591.0  & $BTU/lbm$             & Lower heating value of H2                  \\
        \hline
    \end{tabular}
    \label{engine_params}
\end{table}

@@

\subsection{Problem Statment}
% Briefly discuss the optimization problem
We will perform gradient-based design optimization of the engine model with a multi-point architecture that considers several flight conditions experienced by commercial aircraft.
An XDSM diagram of the optimization problem with the multipoint formulation is shown in Figure \ref{fig:N3_xdsm_opt}.
The objective is to minimize the TSEC at the cruise condition subject to constraints, thrust, max temperature, and engine diameter.
We will perform a parameter sweep of the limiting constraint to understand the impact of different emissions and performance requirements on the design space.
The optimization problem of the N3 engine and the vapor recovery loop are shown in Figure \ref{fig:N3_xdsm_opt}.

@@

% Briefly discuss the optimization problem
We will perform gradient-based design optimization of the engine model with a multi-point architecture that considers several flight conditions experienced by commercial aircraft.
An XDSM diagram of the optimization problem with the multipoint formulation is shown in Figure \ref{fig:opt_prob}.

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=0.75\textwidth]{N3_inject.pdf}
    \caption{XDSM diagram of the multi-point optimization problem with constraints.
        The XDSM diagram shows the variables and outputs within the optimization model and how each of these values is connected to the optimizer and design points.}
    \label{fig:opt_prob}
\end{figure}

The objective is to minimize the TSEC at the cruise condition subject to constraints, thrust, max temperature, and engine diameter.
We will perform a parameter sweep of the limiting constraint to understand the impact of different emissions and performance requirements on the design space.

@@

SNOPT (Sparse Nonlinear OPTimizer) was the optimizer chosen for this work.
SNOPT is a state-of-the-art optimizer used to solve large-scale optimization problems (linear and nonlinear).
The software uses a sparse SQP (sequential quadratic programming) algo- rithm with a limited-memory quasi-Newton approximation to the Hessian of the lagrangian.
SNOPT is particularly useful for nonlinear problems whose functions and derivatives are expensive compute.
An augmented Lagrangian merit function ensures convergence from an arbitrary point.
SNOPT allows the nonlinear constraints to be violated (if necessary) and minimizes the sum of these violations.
Due to the robustness of the algorithm, SNOPT is a perfect candidate to solve the highly complex Multipoint systek since the pyCycle engine cycle would occasionally reach a point that is non-physical and would thus cause the solver to fail.
SNOPT is able to respond to these solver failures and backtrack to a region in the design space that is physical and continue the optimization.

@@

>>>>

% [x] TODO: AL-PA, Consider changing the title to "Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization"
\title{Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization} % change
\author{Peter N. Atma\footnote{MSE Student, Department of Aerospace Engineering, AIAA Student Member}}
\author{Andrew hbt!.R. Lamkin\footnote{Ph.D.~Candidate, Department of Aerospace Engineering, AIAA Student Member}}
\author{Joaquim R.~R.~A.~Martins\footnote{Professor, Department of Aerospace Engineering, AIAA Fellow}}
\affil{University of Michigan, Ann Arbor, MI, 48109}

% ==================================================
%	OVERALL COMMENTS
% ==================================================

@@

For measuring the efficiency of jet engines, thrust-specific fuel consumption (TSFC) is generally used since it is a represents how low the fuel burn is for a given level of thrust.
However, when comparing Jet-A and hydrogen fuels this is not such a good metric since a given mass flow rate of hydrogen has an energy content almost 3 times that of Jet-A.
% TODO: Update mdolab.bib
Therefore, a new metric for comparing the relative efficiencies of engine running on Jet-A versus hydrogen is thrust-specific energy consumption (TSEC) which multiplies TSFC by the lower heating value (LHV) of the fuel %\cite{Adler2023}.

@@

% Update with table from Andrew
\begin{table}[hbt!]
  \centering
  \caption{
    N+3 engine model parameters.
    These parameters specify the engine design used for analysis in this report.
    Atmospheric parameters are presented that specify the flight conditions and humidity of the atmosphere.}
  \begin{tabular}{c c c c}
    \hline
    Parameter             & Value    & Units                 & Comments                                        \\
    \hline
    $W_{TOC}$             & 820.441  & $lbm/s$               & TOC total mass flow rate of air                 \\
    $BPR_{TOC}$           & 23.945   & $-$                   & TOC bypass Ratio                                \\
    $OPR_{TOC}$           & 53.633   & $-$                   & TOC overall pressure ratio                      \\
    $PR_{LPC,TOC}$        & 3.00     & $-$                   & TOC low-pressure compressor pressure ratio      \\
    $PR_{fan,TOC}$        & 1.30     & $-$                   & TOC fan pressure ratio                          \\
    $F_{net,SLS}$         & 28620.84 & $lbf$                 & SLS net thrust                                  \\
    $F_{net,CRZ}$         & 5510.72  & $lbf$                 & CRZ net thrust                                  \\
    $T_{4,TOC}/T_{4,RTO}$ & 0.926    & $-$                   & TOC-to-RTO temperature ratio                    \\
    $T_{4,RTO}$           & 3400.0   & $^{\circ}R$           & RTO combustor temperature                       \\
    $h_{TOC}$             & 0.00017  & $kg_{water}/kg_{air}$ & humidity ratio at TOC \cite{Kalnay1996}         \\
    $h_{CRZ}$             & 0.00017  & $kg_{water}/kg_{air}$ & humidity ratio at CRZ \cite{Kalnay1996}         \\
    $h_{RTO}$             & 0.009    & $kg_{water}/kg_{air}$ & humidity ratio at RTO \cite{Kalnay1996}         \\
    $h_{SLS}$             & 0.009    & $kg_{water}/kg_{air}$ & humidity ratio at SLS \cite{Kalnay1996}         \\
    $LHV_{JetA}$          & 18564.0  & $BTU/lbm$             & Lower heating value of Jet-A \cite{Lawicki2002} \\
    $LHV_{H2}$            & 51591.0  & $BTU/lbm$             & Lower heating value of H2 \cite{toolbox2003}    \\
    \hline
  \end{tabular}
  \label{engine_params}
\end{table}

@@

\subsection{Problem Statment}
% Briefly discuss the optimization problem
% TODO: PA - PA, Condense this and remove repeated parts
We will perform gradient-based design optimization of the engine model with a multi-point architecture that considers several flight conditions experienced by commercial aircraft.
An XDSM diagram of the optimization problem with the multipoint formulation is shown in Figure \ref{fig:N3_xdsm_opt}.
The objective is to minimize the TSEC at the cruise condition subject to constraints, thrust, max temperature, and engine diameter.
We will perform a parameter sweep of the limiting constraint to understand the impact of different emissions and performance requirements on the design space.
The optimization problem of the N3 engine and the vapor recovery loop are shown in Figure \ref{fig:N3_xdsm_opt}.

@@

@@

SNOPT (Sparse Nonlinear OPTimizer) was the optimizer chosen for this work.
SNOPT is a state-of-the-art optimizer used to solve large-scale optimization problems (linear and nonlinear).
The software uses a sparse SQP (sequential quadratic programming) algorithm with a limited-memory quasi-Newton approximation to the Hessian of the lagrangian.
SNOPT is particularly useful for nonlinear problems whose functions and derivatives are expensive compute.
An augmented Lagrangian merit function ensures convergence from an arbitrary point.
SNOPT allows the nonlinear constraints to be violated (if necessary) and minimizes the sum of these violations.
Due to the robustness of the algorithm, SNOPT is a perfect optimizer to solve the highly complex Multipoint systek since the pyCycle engine cycle would occasionally reach a point that is non-physical and would thus cause the solver to fail.
SNOPT is able to respond to these solver failures and backtrack to a region in the design space that is physical and continue the optimization.

@@

\begin{table}[h]
  \centering
  \caption{
  }
  \small
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{r l l l l l}
                & Variable/Function     & Value  & Lower  & Upper & Units           \\
    \toprule
    objective   & $\dot{m}_{fuel,CRZ}$  & _      & -      & -     & $\frac{lbm}{s}$ \\
    \hline
    variables   & $x_{H2O,TOC}$         & 0.053  & 0.0    & -     & -               \\
                & $x_{H2O,RTO}$         & 0.0    & 0.0    & -     & -               \\
                & $x_{H2O,SLS}$         & 0.0    & 0.0    & -     & -               \\
                & $x_{H2O,CRZ}$         & 0.3    & 0.0    & 0.3   & -               \\
                & $PR_{fan,TOC}$        & _      & _      & -     & -               \\
                & $PR_{LPC,TOC}$        & _      & _      & -     & -               \\
                & $PR_{OPR,TOC}$        & _      & _      & -     & -               \\
                & $T_{4,TOC}/T_{4,RTO}$ & _      & _      & -     & -               \\
                & $T_{4,RTO}$           & _      & _      & -     & $^\circ$R       \\
                & $V_{ratio,CRZ}$       & _      & _      & -     &                 \\
    \cline{3-5}
    constraints & $F_{net,TOC}$         & 5800.0 & 5800.0 & -     & lbf             \\
                & $D_{Fan}$             & 100.0  & 100    & -     & $in^2$          \\
    \cline{3-5}
                &                       & Total  &                                  \\
    \bottomrule
  \end{tabular}
  \label{tab:opt_res_JetA}
\end{table}
